---
title: "In-class Exercise 6"
author: "Stefanie Felicia"
date: "4 October, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

```{r}
pacman::p_load(tidyverse, sf, sfdep, tmap, knitr)
```

```{r}
mpsz <- read_rds("data/rds/mpsz_cl.rds")
```

```{r}
busStop <- st_read(dsn = "data/LTADataMall", 
                 layer = "BusStop") %>%
  st_transform(crs = 3414)
```

```{r}
odbus <- read_csv("data/origin_destination_bus_202508.csv")
```

```{r}
tm_shape(mpsz) + tm_polygons(col = "lightgrey", border.col = "white") + tm_shape(busStop) + tm_dots(ol="red", size = 0.1)
```

Since the visualization includes bus stops that are outside of Singapore, we should combine both geometry data and data points using st_joins.

```{r}
busStop_in_SG <- st_join(
  busStop, mpsz, 
  join = st_within, 
  left=FALSE)
```

```{r}
tm_shape(mpsz) + tm_polygons(col = "lightgrey", border.col = "white") + tm_shape(busStop_in_SG) + tm_dots(ol="red", size = 0.1)
```

Next, we'll create grid layer and take hexagon with bus stops and exclude those areas without bus stop (either grid square or hexagon, or even traingle, etc)

```{r}
hexagon <- st_make_grid(mpsz, cellsize = 700, # diameter is 700 from center to edge
                        what = "polygon", # returns hexagon layer
                        square = FALSE) %>%
  st_sf() # make large sfc_polygon data into sf tabular (list) dataframe
```

Since the hexagon table only have 1 column which is the polygon data, we need to assign id to each hexagon to know which it is.

```{r}
hexagon$n_collisions = #return true or false if hexagon intersect the bus stop
  lengths(st_intersects(hexagon, busStop_in_SG))

hexagon <- filter(hexagon, n_collisions > 0) # filter to only store hexagons that has bus stop
```

::: callout-note
-   `hexagon$n_collisions` returns if hexagon intersects the bus stop

-   `lengths(st_intersects(hexagon, busStop_in_SG))` : st_intersections return output of geomteric cut, while this function return true or false (in numerical value; 0 means none or values \> 1 meaning the total number of bus stops within the hexagon),

-   `filter(hexagon, n_collisions > 0)` : filter to only store hexagons that has bus stop
:::

```{r}
hexagon <- hexagon %>%
  select(, -n_collisions)

hexagon$HEX_ID <- sprintf(
  "H%04d", seq_len(nrow(hexagon))) %>% # start with 0
  as.factor() #ordinal scale

kable(head(hexagon))
```

## Preparing Trip Generation Data

### Cleaning Data

```{r}
trips <- odbus %>%
  select(c(ORIGIN_PT_CODE, DAY_TYPE, TIME_PER_HOUR, TOTAL_TRIPS)) %>%
  rename(BUS_STOP_N = ORIGIN_PT_CODE) %>%
  rename(HOUR_OF_DAY = TIME_PER_HOUR) %>%
  rename(TRIPS = TOTAL_TRIPS)
trips$BUS_STOP_N <- as.factor(trips$BUS_STOP_N) # so that it cannot be converted accidentally to numerical column
kable(head(trips))
```

```{r}
# combine bus stop and hexagon (layer) data
bs_hex <- st_intersection(
  busStop, hexagon) %>%
   st_drop_geometry() %>% #geometry will be replaced with hexagon 
  select(c(BUS_STOP_N, HEX_ID))

kable(head(bs_hex))
  
```

### Add HEX_ID into bus trip data

```{r}
trips <- inner_join(trips, bs_hex)
kable(head(trips))
```
