---
title: "In-class Exercise 3b"
author: "Stefanie Felicia"
date: "September 13, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Import library

```{r}
pacman::p_load(tidyverse, sf, httr, tmap)
```

The libraries used in this exercise would be:

-   tidyverse: collection of R packages that for data manipulation and visualization

-   sf: simple features in R to encode and analyze spatial vector data

-   httr: tools to work with URLs and HTTP

-   tmap: used to draw thematic map

## Import data

```{r}
folder_path <- "data/ACRAInformationonCorporateEntities"
file_list <- list.files(path =  folder_path, 
                        pattern = "^ACRA*.*\\csv$", 
                        full.names = TRUE)
acra_data <- file_list %>%
  map_dfr(read_csv)
```

```{r}
write_rds(acra_data, "data/rds/acra_data.rds")
```

## Prepare ACRA data

Tidying the ACRA data include:

-   Select target business using SSIC

-   Derive year and month from registration_incorporation_date

-   Tidy postal code values to ensure avoid 5 digit postal code by mistake

-   Select businesses registered in 2025

```{r}
biz_56111 <- acra_data %>%
  select (1:24) %>%
  filter(primary_ssic_code == 56111) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(regis_date = as.Date(date), 
         YEAR = year(date), 
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date,
                            label = TRUE, 
                            abbr = TRUE)) %>%
  mutate(
    postal_code = str_pad(postal_code, 
                          width = 6, side = "left", pad = "0")) %>%
  filter(YEAR == 2025)
```

## Geocode ACRA data

```{r}
postcodes <- unique(biz_56111$postal_code)
url <- "https://onemap.gov.sg/api/common/elastic/search"
found <- data.frame()
not_found <- data.frame(postcode = character())

for (pc in postcodes){
  query <- list(
    searchVal = pc, 
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )
  res <- GET(url, query = query)
  json <- content(res)
  
  if (json$found != 0){
    df <- as.data.frame(json$results, stringsAsFactors = FALSE)
    df$input_postcode <- pc
    found <- bind_rows(found, df)
  } else {
    not_found <- bind_rows(not_found, data.frame(postcode = pc))
  }
  
}
                        
```

```{r}
biz_56111 = biz_56111 %>%
  left_join(found, 
            by = c('postal_code' = 'POSTAL'))
colnames(biz_56111)
```

```{r}
biz_56111_sf <- st_as_sf(biz_56111, 
                      coords = c("X", "Y"), 
                      crs = 3414)

```
