---
title: "Hands-on Exercise 1a"
author: "Stefanie Felicia"
date: "August 28, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Getting Started

Objective of this exercise:

-   Create thematic map, which involve the use of map symbols to visualize geographic features that are not naturally visible (e.g. population,temperature, crime rate, property prices, etc).

-   Create geovisualization, which uses visual representations and cartographic techniques to explore, analyze and communicate geospatial data

## Import Libraries

```{r}
pacman::p_load(sf, tmap, tidyverse, rvest)
```

The libraries used in this exercise would be:

-   tmap
-   readr: fast way to read rectangular data from delimited files (e.g. csv and tsv)
-   tidyr: tool to create tidy data
-   dplyr: grammar of data manipulation (to work with data frame like objects)
-   sf: simple features in R to encode and analyze spatial vector data
-   rvest: help scrape data from web page

## The Data

In this hands-on exercise, we'll be using the *Master PLan 2019 Master Boundary* (No Sea) KML file (geospatial data), which consist of the geographical boundary of Singapore at planning subzone area in 2019.

We'll also be using the *Singapore Residents by Planning Area/Subzone, Age Group, Sex and Type of Dwelling, June 2024* in csv file (aspatial data). The PA and SZ column can be used to identify to georeference to the *Master Plan 2019 Subzone Boundary* data.

## Import Data

```{r}
mpsz <- st_read("data/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

## Data Preparation
