---
title: "Hands-on Exercise 5a"
author: "Stefanie Felicia"
date: "26 September, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Getting Started

In this hands-on exercise, we'll learn:

-   Convert GIS polygon data to R's sf dataframe

-   Convert simple feature dataframe into R's SpatialPolgonDataFrame object

-   Perform cluster analysis by using hclust()

-   Perform spatially constrained cluster analysis using skater()

-   Visualize nalaysis output using ggplot2 and tmap

In this hands-on exercise, we'll learn how to delineate the market or planning area to homogeneous regions by using multivariate data. We'll delinate Shan State, Myanmar unto homogeneous regions by using multiple information and Communication Technology measures (i.e. radio, TV, land line phone, mobile phone, computer adn internet at home)

## Import Libraries

```{r}
pacman::p_load(spdep, tmap, sf, ClustGeo, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse, GGally)
```

The libraries used in this exercise would be:

-   sf: simple features in R to encode and analyze spatial vector data

-   spdep: create spatial weights matrix object from polygon 'contiguities'

-   tmap: used to draw thematic map

-   tidyverse: collection of R packages that for data manipulation and visualization

-   GGally: extend ggplot2 by adding several functions to reduce complexity of combining geometric object with transformed data

-   psych: general purpose toolbox developed originally for personality, psychometric theory and experimental psychology

-   corrplot: exploratory tool for correlation matrix that support variable reordering to help detect hidden pattern (by visualizing the lower and upper bound of confidence interval

-   heatmaply: create interactive cluster 'heatmaps' that can be saved as HTMl file, Shiny app or embedded in R markdown

-   factoextra: extract and visualize the output of multivariate data analyses

-   NbClust: provide 30 indexes for determining the optimal number of cluster in data set

-   ggpubr: create beautiful ggplot2-based graphs that is easy to use

-   cluster: find groups in data (cluster analysis)

-   ClustGeo: implement a Ward-like hierarchical clustering algorithm(incl. spatial/geographical constraints)

## The Data

In this hands-on we'll be using these dataset:

-   Myanmar Township Boundary Data (myanmar_township_boundaries) which is a GIS data in ESRI shapefile.

-   Shan-ICT.csv from The 2014 Myanmar Population and Housing Census Myanmar

## Import Data

We'll import the boundary data using st_read()

```{r}
shan_sf <- st_read(dsn = "data/geospatial", 
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>%
  select(c(2:7))
```

```{r}
shan_sf
```

As can be seen, the sf dataframe is conformed to Hardy Wickham's tidy framework. We can confirm this by using glimpse() to check the data type of the fields.

```{r}
glimpse(shan_sf)
```

Next, we'll import the csv file with read_csv() function

```{r}
ict <- read_csv("data/aspatial/Shan-ICT.csv")
```

```{r}
summary(ict)
```

## Data Preparation

Since the unit measurement of the values are number of household, it can create bias by the underlying total number of household. Townships with higher number of household will also have higher number of household that own radio, TV, etc. So we'll derive penetration rate of each ICT variable.

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 
```

```{r}
summary(ict_derived)
```

New columns derived are RADIO_PR, TV_PR, LLPHONE_PR, MPHONE_PR, COMPUTER_PR and INTERNET_PR

## EDA

Next, let's plot the distribution the variable of number of household with radio.

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

We'll also visualize outliers with boxplot

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

Next, we'll visualize the radio penetration rate

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

We'll also visualize other newly derived variables

```{r}
radio <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

llphone <- ggplot(data=ict_derived, 
             aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

mphone <- ggplot(data=ict_derived, 
             aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

computer <- ggplot(data=ict_derived, 
             aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

internet <- ggplot(data=ict_derived, 
             aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r}
ggarrange(radio, tv, llphone, mphone, computer, internet, 
          ncol = 3, 
          nrow = 2)
```

We can also understand better the by visualizing using choropleth. But before visualizing we need to join the geospatial and aspatial data by using left_join()

```{r}
shan_sf <- left_join(shan_sf, 
                     ict_derived, by=c("TS_PCODE"="TS_PCODE"))
  
write_rds(shan_sf, "data/rds/shan_sf.rds")
```

```{r}
shan_sf <- read_rds("data/rds/shan_sf.rds")
```

Now, we can visualize distribution of RADIO_PR

```{r}
qtm(shan_sf, "RADIO_PR")
```

The choropleth shows that the distribution is baised to the total number of household at townships, so we'll create two choropleth (number of household and number of household with radio) to compare.

```{r}
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) + 
  tm_fill(col = "TT_HOUSEHOLDS",
          n = 5,
          style = "jenks", 
          title = "Total households") + 
  tm_borders(alpha = 0.5) 

RADIO.map <- tm_shape(shan_sf) + 
  tm_fill(col = "RADIO",
          n = 5,
          style = "jenks",
          title = "Number Radio ") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map,
             asp=NA, ncol=2)
```

As observed, the townships that has higher number of household have higher number of radio. Next, let's plot the distribution of total number of household and radio penetration rate.

```{r}
tm_shape(shan_sf) +
    tm_polygons(c("TT_HOUSEHOLDS", "RADIO_PR"),
                style="jenks") +
    tm_facets(sync = TRUE, ncol = 2) +
  tm_legend(legend.position = c("right", "bottom"))+
  tm_layout(outer.margins=0, asp=0)
```

As can be seen the townships where there is penetration of radio are not townships that has higher households.

## Correlation Analysis

Before performing clustering, we need to check correlation of each variable, since the requirements of clustering is that varaibles with high correlations shouldn't be included in clustering ( \> 0.8 correlation).

```{r}
cluster_vars.cor = cor(ict_derived[,12:17])
corrplot.mixed(cluster_vars.cor,
         lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

Since COMPUTER_PR and INTERNET_PR are highly correlated, only one of them should be used

## Hierarchy Clustering Analysis

These are the steps we'll be doing for performing hierarchical clusetring analysis

1.  Extract variables to be used for clustering
2.  Change row number into township name
3.  Data standardization
4.  Compute proximity matrix
5.  Select optimal clustering algorithm
6.  Determine optimal cluster number
7.  Interpret results

### Extract Variables to be Used for Clustering

```{r}
cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>%
  select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
head(cluster_vars,10)
```

We decided to not include INTERNET_PR because it's highly correlated with COMPUTER_PR.

### Change Row Number into Township Name

```{r}
row.names(cluster_vars) <- cluster_vars$"TS.x"
head(cluster_vars,10)
```

Next, we will delete the TS.x field

```{r}
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 10)
```

```{r}
write_rds(shan_ict, "data/rds/shan_ict.rds")
```

```{r}
shan_ict <- read_rds("data/rds/shan_ict.rds")
```

### Data standardization

Fields used cluster analysis usually has different value ranges, so to avoid biased results, we usually standardize the input variable first. There are different standardization methods: min-max and z-score

#### Min-Max Standardization

We can use the normalize() of heatmaply library to standardize using Min-Max methods.

Note: min–max standardization is a technique to transform features within a certain range (usually 0 to 1)

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

#### Z-score Standardization

Z-score standardization is performed usually with scale() like below

Note: z-score can only be used when the feature has approximately normal distribution

```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

#### Visualize Standardized Clustering Variables

It's a good practice to check the distribution of the standardized variables. In this example, we'll look at RADIO_PR field

```{r}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

As can be seen, standardization doesn't change the distribution of the feature (except for the count of 0.5). It can also be seen that the min-max and z-score standardization doesn't seem much different from each other.

```{r}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

### Compute Proximity Matrix

We can calculate proximity matrix using dist(). R also supports different proximity calculations: euclidean, maximum, manhattan, canberra, binary and minkowski. But default used in this function is euclidean method

```{r}
proxmat <- dist(shan_ict, method = 'euclidean')
```

```{r}
proxmat
```

### Compute Hierarchical Clustering

In this hands-on exercise, we'll perform hierarchical clustering with hclust(), but there are other clustering functions avaiable.

There are a few algorithm to compute cluster such as ward.D, ward.D2, single, complete, average (UPGMA), mcquitty (WPGMA), median (WPGMC) and centroid (UPGMC). In this example, we'll use ward.D method.

```{r}
hclust_ward <- hclust(proxmat, method = 'ward.D')
```

Next, we can print the dendogram like so.

```{r}
plot(hclust_ward, cex = 0.6)
```

### Select Optimal Clustering Algorithm

The challenge in performing hierarchical clustering is to identify stronger clustering structures, but it can be resolved with agnes() function from cluster library. The function not only perform hierarchical clustering, but also provide agglomerative coefficient (measuring amount of clustering structure found, so the values closer to 1 means strong clustering structure)

```{r}
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

ac <- function(x) {
  agnes(shan_ict, method = x)$ac
}

map_dbl(m, ac)
```

As can be seen, the Ward's method has strongest clustering structure among 4 methods. So we'll be using Ward's method.

### Determining Optimal Clusters

Another challenge faced when performing clustering analysis is to determine the optimal number of clusters. The commonly used methods are the following: Elbow Method, Average Silhouette Method and Gap Statistic Method.

#### Gap Statistic Method

This function compares the total within intra cluster variation for different values if k with expected values under null reference distribution of the data. The estimate of the optimal clusters will be the value that maximize the gap statistics (yield the largest gap statistics), meaning the clustering structure is far from random distribution of points. clusGap() of cluster library is used to compute gap statistics.

```{r}
set.seed(12345)
gap_stat <- clusGap(shan_ict, 
                    FUN = hcut, 
                    nstart = 25, 
                    K.max = 10, 
                    B = 50)
# Print the result
print(gap_stat, method = "firstmax")
```

We can visualize it using fvix_gap_stat().

```{r}
fviz_gap_stat(gap_stat)
```

As can be seen, 6-cluster gives largest gap statistics and should be the next best cluster to pick.

### Interpret Dendograms

We can interpret dendogram by looking at each leaf, as they correspond to one observation. As we move up, observations that are similar will be combined into a branch. The higher we move up, the more different they are from each other.

We can draw dendogram with a border around the selected clusters using rect.hclust()

```{r}
plot(hclust_ward, cex = 0.6)
rect.hclust(hclust_ward, 
            k = 6, 
            border = 2:5)
```

### Visually Driven Hierarchical Clustering Analysis

IN this section, we'll perform visaully-driven hierrachical clustering analysis using heatmaply().

Before plotting the interactive cluster, we;ll transform the data frame into a matrix to create the heatmap.

```{r}
shan_ict_mat <- data.matrix(shan_ict)
```

Next, we'll use heatmpaly() to build an interactive cluster heatmap.

```{r}
heatmaply(normalize(shan_ict_mat),
          Colv=NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State"
          )
```

### Map Clusters Formed

Seeing the dendogram, we decide to use 6 clusters. We'll be using cutree() to derive 8 cluster model.

```{r}
groups <- as.factor(cutree(hclust_ward, k=6))
```

To visualize the clusters, the groups object need to be appended to shan_sf.

```{r}
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER`=`as.matrix.groups.`)
```

::: callout-note
-   The groups list object will be converted into a matrix

-   cbind() is used to append groups matrix to shan_sf to produce output sf object called shan_sf_cluster

-   rename as.matrix.groups field as CLUSTER
:::

Next, we can plot the choropleth map

```{r}
qtm(shan_sf_cluster, "CLUSTER")
```

As can be seen, the clusters are fragmented and this is the limitation when we use non-spatial clustering to do spatial clustering.

## SKATER Approach

SKATER approach is one of the alternative to do clustering on spatial data. It is best to be used when we have continuous or multivariate data.

This approach works by creating connectivity graph between neighbouring area where each area is represented by node and edges to represent connection between area.

Step by step on how to use SKATER approach:

1.  Convert sf object to SpatialPolygonsDataframe
2.  Compute neighbour list
3.  Calculate edge cost
4.  Compute minimum spanning tree
5.  Compute spatially constrained cluster with SKATER method (when to stop, it needs to be n-1 edges that would not split graph into disconnecting parts). Nodes with higher dissimilarity cost will be removed

### 1. Convert sf object to SpatialPolygonsDataframe

Like before, we'll convert sf data to SpatialPolygonDataframe since SKATER function only accepts sp objects

```{r}
shan_sp <- as_Spatial(shan_sf)
```

### 2. Compute Neighbour List

Next, we'' compute the neighbour list from polygon list using poly2nd()

```{r}
shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

We can plot the neighbour_list and its boundaries, by plotting the graph on top of the map.

```{r}
coords <- st_coordinates(
  st_centroid(st_geometry(shan_sf)))
```

```{r}
plot(st_geometry(shan_sf), 
     border=grey(.5))
plot(shan.nb,
     coords, 
     col="blue", 
     add=TRUE)
```

::: callout-note
-   `plot(geometry(...)`map the boundary

-   `plot(shan.nb...)` map the neighbour list and coordinate applied to original sp daframe ro extract the centroid of the polygons (as graph representation)
:::

### 3. Calculate Edge Cost

We'll use nbcosts() to compute the cost of each edge (distance between each nodes). So for each observation, it gives pairwise dissimilarity between its values on the five variables and values for the neighbouring observation (from neighbours list).

```{r}
lcosts <- nbcosts(shan.nb, shan_ict)
```

Next, we'll apply the cost into weights object as we did in the calculation of inverse distance weights. We'll use nb2listw() and use `style = "B"` to make sure the cost values are not row-standardized.

```{r}
shan.w <- nb2listw(shan.nb, 
                   lcosts, 
                   style="B")
summary(shan.w)
```

### 4. Compute Minimum Spanning Tree

We'll compute the minimum spanning tree using mstree()

```{r}
shan.mst <- mstree(shan.w)
```

```{r}
# check class
class(shan.mst)
```

```{r}
# check dimension (no of rows and columns)
dim(shan.mst)
```

The dimension is 54 because minimum spanning tree consist of n-1 edge to traverse all nodes. we can also check for content of minimum spanning tree as below.

```{r}
head(shan.mst)
```

We can also visualize it like so.

```{r}
plot(st_geometry(shan_sf), 
                 border=gray(.5))
plot.mst(shan.mst, 
         coords, 
         col="blue", 
         cex.lab=0.7, 
         cex.circles=0.005, 
         add=TRUE)
```

### 5. Compute Minimum Spanning Tree

```{r}
clust6 <- spdep::skater(edges = shan.mst[,1:2], 
                 data = shan_ict, 
                 method = "euclidean", 
                 ncuts = 5)
```

We used the skater() function which takes 3 arguments:

-   `edges =` the first two columns of the MST matrix (i.e. not the cost)

-   `data =` is the data matrix (to update the cost as units are being grouped

-   `ncuts =` it's one less than the number of cluster. So the value specified is not the number of clusters, but the number of cuts in the graph, one less than the number of clusters.

```{r}
str(clust6)
```

The most interesting component of this list is the groups vector containing the labels of the cluster to which each observation belongs. This is followed by a detailed summary for each of the clusters in the edges.group list. Sum fo squares measures are given as for the total and ssw to show the effect of each of the cuts on the overall criterion.

We can check the cluster assignment like so

```{r}
ccs6 <- clust6$groups
ccs6
```

We can find out how many observation are in each cluster by means of the table command. We can also find this as the dimension of each vector in the lists contained in edges.groups (e.g. the first list has node with dimension 12, which is also the number of observation in the first cluster.

```{r}
table(ccs6)
```

We can also plot pruned tree that shows the five clusters on top of the townshop srea

```{r}
plot(st_geometry(shan_sf), 
     border=gray(.5))
plot(clust6, 
     coords, 
     cex.lab=.7,
     groups.colors=c("red","green","blue", "brown", "pink"),
     cex.circles=0.005, 
     add=TRUE)
```

Lastly we can also plot the newly derived clusters by using SKATER method

```{r}
groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename(`SP_CLUSTER`=`as.factor.groups_mat.`)
qtm(shan_sf_spatialcluster, "SP_CLUSTER")
```

And side by side comparison

```{r}
hclust.map <- qtm(shan_sf_cluster,
                  "CLUSTER") + 
  tm_borders(alpha = 0.5) 

shclust.map <- qtm(shan_sf_spatialcluster,
                   "SP_CLUSTER") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(hclust.map, shclust.map,
             asp=NA, ncol=2)
```
