---
title: "In-class_Ex07"
format: html
---

```{r}
pacman::p_load(olsrr, corrplot, ggpubr, 
               sf, sfdep, GWmodel, tmap, 
               tidyverse, gtsummary,
               performance, RColorBrewer)
```

The libraries used in this exercise would be:

-   olsrr

-   ggpubr

-   sf

-   sfdep

-   GWmodel

-   tidyverse

-   tmap

-   gtsummary

-   performance

-   RColorBrewer

## The Data

We'll be using the following dataset for this hands on exercise:

-   URA Mater Plan Subzone boundary (shapefile format)

-   condo_resale_2015 (csv file)

## Import Data

We'll import the shapefile dataset with st_read(). The dataset consist of URA Master Plan 2014 planning subzone boundaries and it's in svy21 projected coordinate systems.

```{r}
mpsz = st_read(dsn = "data/geospatial", 
               layer = "MP14_SUBZONE_WEB_PL")
```

```{r}
mpsz_svy21 <- st_transform(mpsz, 3414)
```

```{r}
st_crs(mpsz_svy21)
```

```{r}
condo_resale = read_csv(
  "data/aspatial/Condo_resale_2015.csv")
```

```{r}
condo_resale.sf <- st_as_sf(condo_resale,
                            coords = c("LONGITUDE", "LATITUDE"),
                            crs=4326) %>%
  st_transform(crs=3414)
```

```{r}
condo_resale.sf <- condo_resale.sf %>%
  mutate(`LOG_SELLING_PRICE` = log(SELLING_PRICE))
names(condo_resale.sf)
```

```{r}
condo.mlr1 <- lm(
  formula = SELLING_PRICE ~ AREA_SQM + AGE + 
    PROX_CBD + PROX_CHILDCARE + PROX_MRT +
    PROX_ELDERLYCARE    + PROX_URA_GROWTH_AREA +
    PROX_PARK   + PROX_PRIMARY_SCH + 
    PROX_SHOPPING_MALL  + PROX_BUS_STOP + 
    NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD,
  data=condo_resale.sf)
summary(condo.mlr1)
```


```{r}
mlr.output <- as.data.frame(condo.mlr1$residuals)
```



```{r}
names(condo_resale.sf)
```

```{r}
condo_resale.sf <- cbind(condo_resale.sf, 
                        condo.mlr1$residuals)  %>%
rename(`MLR_RES` = `condo.mlr1.residuals`)

names(condo_resale.sf)
```

```{r}
condo_resale.sf <- condo_resale.sf %>% 
  mutate(
    nb = st_dist_band(
      st_geometry(geometry), 
                 upper = 1500),
    wt = st_weights(
      nb, style = "W"),
    .before = 1) 

```

```{r}
set.seed(1234)
global_moran_perm(
  condo_resale.sf$MLR_RES, 
  nb = condo_resale.sf$nb, 
  wt = condo_resale.sf$wt,
  alternative = "two.sided", 
  nsim = 499)
```

```{r}
bw.adaptive <- bw.gwr(
  formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE    +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL   + 
    PROX_BUS_STOP   + NO_Of_UNITS + 
    FAMILY_FRIENDLY + FREEHOLD, 
  data=condo_resale.sf, 
  approach="AIC", 
  kernel="gaussian", 
  adaptive=TRUE, 
  longlat=FALSE)
```

```{r}
gwr.adaptive <- gwr.basic(
  formula = SELLING_PRICE ~ AREA_SQM + AGE + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE +
    PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_SHOPPING_MALL + PROX_BUS_STOP +
    NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD,
  data=condo_resale.sf, 
  bw=bw.adaptive, 
  kernel = 'gaussian', 
  adaptive=TRUE, 
  longlat = FALSE)
```

```{r}
gwr.adaptive
```

CONVERT SDF INTO SF DATA FRAME

```{r}
condo_resale.sf.adaptive <-
  st_as_sf(gwr.adaptive$SDF) %>%
  st_transform(crs=3414)
```

```{r}
glimpse(condo_resale.sf.adaptive)
```

```{r}
tmap_mode("plot")
tm_shape(mpsz_svy21)+
  tm_polygons(fill_alpha = 0.1) +
tm_shape(condo_resale.sf.adaptive) +  
  tm_dots(col = "Local_R2",
          border.col = "gray60",
          border.lwd = 1) +
  tm_view(set.zoom.limits = c(11,14))


```

```{r}
AREA_SQM_SE <- tm_shape(mpsz_svy21)+
  tm_polygons(alpha = 0.1) +
tm_shape(condo_resale.sf.adaptive) +  
  tm_dots(col = "AREA_SQM_SE",
          border.col = "gray60",
          border.lwd = 1) +
  tm_view(set.zoom.limits = c(11,14))

AREA_SQM_TV <- tm_shape(mpsz_svy21)+
  tm_polygons(alpha = 0.1) +
tm_shape(condo_resale.sf.adaptive) +  
  tm_dots(col = "AREA_SQM_TV",
          border.col = "gray60",
          border.lwd = 1) +
  tm_view(set.zoom.limits = c(11,14))

tmap_arrange(AREA_SQM_SE, AREA_SQM_TV, 
             asp=1, ncol=2,
             sync = TRUE)

tmap_mode("plot")
```

## Analyzing R² Distribution in Novena Planning Area


### Extracting Novena Data


# Extract Novena planning area boundary


```{r}
novena_boundary <- mpsz_svy21 %>% 
  filter(PLN_AREA_N == "NOVENA")
```

# Filter GWR results for Novena area

```{r}
novena_gwr <- condo_resale.sf.adaptive %>% 
  st_join(novena_boundary) %>% 
  filter(!is.na(PLN_AREA_N))

```

# Check the number of observations in Novena

```{r}
nrow(novena_gwr)
```

### Summary Statistics of R² in Novena


# Summary of Local R² in Novena

```{r}
novena_r2_summary <- summary(novena_gwr$Local_R2)
novena_r2_summary
```
# Additional statistics

```{r}
novena_r2_stats <- novena_gwr %>% 
  st_drop_geometry() %>% 
  summarise(
    n = n(),
    mean_r2 = mean(Local_R2),
    median_r2 = median(Local_R2),
    sd_r2 = sd(Local_R2),
    min_r2 = min(Local_R2),
    max_r2 = max(Local_R2),
    q1_r2 = quantile(Local_R2, 0.25),
    q3_r2 = quantile(Local_R2, 0.75)
  )
novena_r2_stats
```

### Visualizing R² Distribution in Novena


# Histogram of Local R² in Novena

```{r}
ggplot(novena_gwr, aes(x = Local_R2)) +
  geom_histogram(bins = 15, fill = "lightblue", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = mean(Local_R2)), color = "red", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Local R² in Novena Planning Area",
       x = "Local R²",
       y = "Frequency") +
  theme_minimal()
```

### Spatial Distribution of R² in Novena


# Map of Local R² in Novena
```{r}
tm_shape(novena_boundary) +
  tm_polygons(alpha = 0.3, col = "lightgray") +
tm_shape(novena_gwr) +
  tm_dots(col = "Local_R2", 
          palette = "RdYlBu",
          style = "quantile",
          size = 0.8,
          title = "Local R²") +
  tm_layout(main.title = "Local R² Distribution in Bishan Planning Area",
            legend.outside = TRUE)
```


### Comparison with Overall R² Distribution


# Compare Novena with overall distribution

```{r}
overall_r2_stats <- condo_resale.sf.adaptive %>% 
  st_drop_geometry() %>% 
  summarise(
    n = n(),
    mean_r2 = mean(Local_R2),
    median_r2 = median(Local_R2),
    sd_r2 = sd(Local_R2)
  )
```

```{r}
comparison <- data.frame(
  Area = c("Novena overall"),
  Mean_R2 = c(novena_r2_stats$mean_r2, overall_r2_stats$mean_r2),
  Median_R2 = c(novena_r2_stats$median_r2, overall_r2_stats$median_r2),
  SD_R2 = c(novena_r2_stats$sd_r2, overall_r2_stats$sd_r2)
)
comparison
```
 