---
title: "Hands-on Exercise 8"
author: "Stefanie Felicia"
date: "25 October, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Getting Started

Predictive model use statistical learning or machine learning techniques to predict its outcome in the future using relevant variables.

Geospatial predictive modeling is conceptually rooted in the principle that the occurrences of events being modeled are limited in distribution. When geographically referenced data are used, occurrences of events are neither uniform nor random in distribution over space. There are geospatial factors (infrastructure, sociocultural, topographic, etc) that constrain and influence where the location of events occur. Geospatial predictive modeling attempt to describe those constraints and influences by spatially correlating occurrences of historical geospatial locations with environmental factors. So in this hands on exercise, we'll learn to build predictive model using geographical random frorest method.

## Import Libraries

```{r}
pacman::p_load(sf, spdep, GWmodel, SpatialML, 
               tmap, rsample, Metrics, tidyverse)
```

The libraries used in this exercise are:

-   sf: simple features in R to encode and analyze spatial vector data

-   spdep: create spatial weights matrix object from polygon 'contiguities'

-   tmap: used to draw thematic map

-   tidyverse: collection of R packages that for data manipulation and visualization

-   GWmodel: techniques form particulae branch of spatial statistics termed geographically weighted (GW model)

-   SpatialML: implement spatial extension of random forest algorithm

-   rsample: classes adn functions to create and summarize different type sof resampling objects (e.g. bootstrap, cross-validation)

-   Metrics: evaluation metrics in R for supervised ML

## The Data

The data we'll be using in this hands on exercise are:

-   HDB Resale data: list of HDB resale transaction prices in Singapore from Jan 2017 onwards

-   MP14_SUBZONE_WEB_PL: polygon feature data providing information of URA 2014 Master Plan Planning Subzone Boundary data (ESRI shapefile)

-   Location factors with geographic coordinates:

    -   Eldercare data: list of eldercare in Singapore (from data.gov.sg)

    -   Hawker centre: list of hawker centre in Singapore (from data.gov.sg)

    -   Supermarket: list of supermarket in Singapore (from data.gov.sg)

    -   Park: list of parks in Singapore (from data.gov.sg)

    -   CHAS Clinic: list of CHAS clinic in Singapore (from data.gov.sg)

    -   Childcare Service: list of childcare service in Singapore (from data.gov.sg)

    -   Kindergatens: list of kindergarten in Singapore (from data.gov.sg)

    -   MRT data: list of MRT/LRT stations in Singapore with station name and codes (from datamall.lta.gov.sg)

    -   Bus Stop: list of bus stops in Singapore (from datamall.lta.gov.sg)

-   Locational factors without geographic coordinates:

    -   Primary school: data extracted from the list on General Information of schools

    -   CBD coordinates (from Google)

    -   Shopping malls: list of shopping amlls in Singapore (from Wikipedia)

    -   Good primary schools: list of primary schools ordered by popularity ranking (from local Salary Forum)

## Data Preparation

```{r}
mdata <- read_rds("data/rawdata/mdata.rds")
```

### Data Sampling

Next, we'll split the dataset 65% and 35% for training and testing using initial_split()

```{r}
set.seed(1234)
resale_split <- initial_split(mdata, 
                              prop = 6.5/10,)
train_data <- training(resale_split)
test_data <- testing(resale_split)
```

The output will be saved into rds file.

```{r}
write_rds(train_data, "data/model/train_data.rds")
write_rds(test_data, "data/model/test_data.rds")
```

### Derive Correlation Matrix

Before we create the model, its good practice to use correlation matrix to examine signs of multicollinearity.

```{r}
mdata_nogeo <- mdata %>%
  st_drop_geometry()
corrplot::corrplot(cor(mdata_nogeo[, 2:17]), 
                   diag = FALSE, 
                   order = "AOE",
                   tl.pos = "td", 
                   tl.cex = 0.5, 
                   method = "number", 
                   type = "upper")
```

As can be seen, all of the values are less than 0.8, which means that the variables are not strongly correlated.

### Retrieve Stored Data

```{r}
train_data <- read_rds("data/model/train_data.rds")
test_data <- read_rds("data/model/test_data.rds")
names(train_data)
```

## Build Non-Spatial Multiple Linear Regression

```{r, eval=FALSE}
price_mlr <- lm(resale_price ~ floor_area_sqm +
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                data=train_data)
summary(price_mlr)
```

As can be seen, the model performs well achieving the score of 0.737 on adjusted R-square and is significant (p-value less thna 0.05). Apart from that, all of the variables used for prediction are contributing to teh model.

We'll save the results in rds file.

```{r, eval=FALSE}
write_rds(price_mlr, "data/model/price_mlr.rds" ) 
```

## gwr Predictive Model

In this section, we'll learn hwo to calibrate model to predict HDB resale price using geographically weighted regression.

### Compute Adaptive Bandwidth

First, we;ll use bw.gwr() to determine optimal bandwidth to be used. We'll be using CV method to determine optimal bandwidth

```{r, eval=FALSE}
bw_adaptive <- bw.gwr(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths +
    PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
    PROX_MRT + PROX_PARK + PROX_MALL + 
    PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
    WITHIN_1KM_PRISCH,
  data=train_data,
  approach="CV",
  kernel="gaussian",
  adaptive=TRUE,
  longlat=FALSE)

write_rds(bw_adaptive, "data/model/bw_adaptive.rds")
```

The output shows that 40 neighbour points will be the optimal bandwidth to be used if adaptive bandwidth is used for the data set.

### Construct Adaptive Bandwidth gwr Model

We'll reload the bandwidth like below

```{r}
bw_adaptive <- read_rds(
  "data/model/bw_adaptive.rds")
```

Now we'll calibrate the gwr-based hedonic pricing model using adaptive bandwidth and Gaussian kernel

```{r, eval=FALSE}
gwr_adaptive <- gwr.basic(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths + 
    PROX_CBD + PROX_ELDERLYCARE + 
    PROX_HAWKER + PROX_MRT + PROX_PARK + 
    PROX_MALL + PROX_SUPERMARKET +
    WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + 
    WITHIN_350M_BUS + WITHIN_1KM_PRISCH,
  data=train_data,
  bw=bw_adaptive, 
  kernel = 'gaussian', 
  adaptive=TRUE,
  longlat = FALSE)

write_rds(gwr_adaptive, "data/model/gwr_adaptive.rds")
```

### Retrieve gwr Output Object

We'll load the saved gwr model

```{r}
gwr_adaptive <- read_rds("data/model/gwr_adaptive.rds")
```

```{r}
gwr_adaptive
```

gwr_adaptive performed better than the non-spatial multiple regression model as its adjusted r square is at 0.96, which means that the model can explain 96% of the resale price. The AICC fits better with the gwr_adaptive.

### Compute Adaptive Bandwidth for Test Dara

```{r}
gwr_bw_test_adaptive <- bw.gwr(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths +
    PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
    PROX_MRT + PROX_PARK + PROX_MALL + 
    PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
    WITHIN_1KM_PRISCH,
  data=test_data,
  approach="CV",
  kernel="gaussian",
  adaptive=TRUE,
  longlat=FALSE)

write_rds(gwr_bw_test_adaptive, 
          "data/model/gwr_bw_test_adaptive.rds")
```

### Compute Predicted Values of Test Data

```{r}
coords <- st_coordinates(train_data)  # extract longitude & latitude (or x/y)
train_data$x <- coords[, 1]
train_data$y <- coords[, 2]
names(train_data)
```

```{r}
coords_test <- st_coordinates(test_data)  # extract longitude & latitude (or x/y)
test_data$x <- coords_test[, 1]
test_data$y <- coords_test[, 2]
names(train_data)
```

```{r, eval=FALSE}
gwr_pred <- gwr.predict(
  formula = resale_price ~ floor_area_sqm + 
    storey_order + remaining_lease_mths + 
    PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER + 
    PROX_MRT + PROX_PARK + PROX_MALL + 
    PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + WITHIN_350M_BUS + 
    WITHIN_1KM_PRISCH, 
  data=train_data, 
  predictdata = test_data, 
  bw=40, 
  kernel = 'gaussian', 
  adaptive=TRUE, 
  longlat = FALSE)

write_rds(gwr_pred, 
          "data/model/gwr_pred.rds")
```

## Calibrate Random Forest Model

### Prepare Coordinate Data

Before we create the random forest model, we'll extract the x and y coordinate for train data and test data

```{r}
coords <- st_coordinates(mdata)
coords_train <- st_coordinates(train_data)
coords_test <- st_coordinates(test_data)
```

We'll save the output in rds file.

```{r}
coords_train <- write_rds(
  coords_train, 
  "data/model/coords_train.rds" )
coords_test <- write_rds(
  coords_test, 
  "data/model/coords_test.rds" )
```

### Drop Geometry Field

Next, we'll drop geometry field using st_drop_geometry()

```{r}
train_data_nogeom <- train_data %>% 
  st_drop_geometry()

test_data_nogeom <- test_data %>% 
  st_drop_geometry()
```

### Calibrate a Non-spatial Random Forest Model

```{r}
set.seed(1234)
rf <- ranger(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths + 
    PROX_CBD + PROX_ELDERLYCARE + 
    PROX_HAWKER + PROX_MRT + PROX_PARK + 
    PROX_MALL + PROX_SUPERMARKET +
    WITHIN_350M_KINDERGARTEN + 
    WITHIN_350M_CHILDCARE + 
    WITHIN_350M_BUS +  
    WITHIN_1KM_PRISCH,
  data=train_data_nogeom)
rf
```

Next, we'll save output into rds file.

```{r}
write_rds(rf, "data/model/rf.rds")
```

### Calibrate Geographical Weighted Random Forest Model

In this section, we'll create weighted random forest model using grf()

#### Create Model Using Training Data

```{r}
set.seed(1234)
gwRF_adaptive <- grf(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths + 
    PROX_CBD + PROX_ELDERLYCARE + 
    PROX_HAWKER + PROX_MRT + PROX_PARK + 
    PROX_MALL + PROX_SUPERMARKET +
    WITHIN_350M_KINDERGARTEN +
    WITHIN_350M_CHILDCARE + 
    WITHIN_350M_BUS + WITHIN_1KM_PRISCH,
  dframe=train_data_nogeom, 
  bw=40,
  kernel="adaptive",
  coords=coords_train)

write_rds(gwRF_adaptive, "data/model/gwRF_adaptive.rds")
```

```{r}
gwRF_adaptive <- read_rds("data/model/gwRF_adaptive.rds")
```

#### Predict Using Test Data

Well prepare tthe test data by combining the data with its coordinates

```{r}
test_data_nogeom <- cbind(
  test_data, coords_test) %>%
  st_drop_geometry()
```

We'll use predict.grf() to predict resale value by using tets data and gwRF_adaptive model.

```{r}
gwRF_pred <- predict.grf(gwRF_adaptive, 
                           test_data_nogeom, 
                           x.var.name="X",
                           y.var.name="Y", 
                           local.w=1,
                           global.w=0)

GRF_pred <- write_rds(gwRF_pred, "data/model/GRF_pred.rds")
```

#### Convert Predicting Output to Dataframe

Since the output of predict.grf() is in vector form, we'll convert it to dataframe for easy visualization and analysis.

```{r}
GRF_pred <- read_rds(
  "data/model/GRF_pred.rds")
GRF_pred_df <- as.data.frame(GRF_pred)
```

We use cbind() to append the predicted valyes to test_data_p

```{r}
test_data_p <- cbind(
  test_data, GRF_pred_df) %>%
  select(resale_price, GRF_pred)
```

```{r}
write_rds(test_data_p, 
          "data/model/test_data_p.rds")
```

### Calculate Root Mean Square Error

Next, we'll evaluate the performance of the model using root mean square error. This works by meausring how far predicted values are from observed values in regression analysis.

```{r}
rmse(test_data_p$resale_price, 
     test_data_p$GRF_pred)
```

Next, we'll visualize the predicted value.

```{r}
ggplot(data = test_data_p,
       aes(x = GRF_pred,
           y = resale_price)) +
  geom_point()
```
