---
title: "Hands-on Exercise 3"
author: "Stefanie Felicia"
date: "September 11, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Getting Started

Spatio-temporal point process or also known as space time point process, is a random collection of point that represents time and location of an event (e.g. disease outbreak, sighting of a species, lightning sightings, etc). The need for spatio-temporal point pattern has become increasingly vital as there has been rapid emergence of geographically ad temporally indexed data in a lot of fields.

In this exercise, we'll explore how to run a set of spatio-temporal point pattern analysis in R by using the forest fire event in Bangka Belitung, Indonesia. We'll be specifically answering these questions:

-   Are locations of forest fire in Bangka Belitung spatial and spatio-temporally independent ?

-   If not, where and when does the observed forest fire tend to cluster ?

## Import Libraries

```{r}
pacman::p_load(sf, raster, spatstat, sparr, tmap, tidyverse)
```

The libraries used in this exercise would be:

-   sf: simple features in R to encode and analyze spatial vector data

-   raster: read, write, manipulate, analyze and model spatial data

-   spatstat: spatial statistics with focus on analyzing spatial point patterns in SD (with some 3D support)

-   sparr: provide function to estimate kernel-smoothed spatial and spatio-temporal densities and relative risk function and perform subsequent inference

-   tmap: used to draw thematic map

-   tidyverse: collection of R packages that for data manipulation and visualization

## The Data

In this hands-on exercise, we'll be using the following datasets:

-   forestfires: a csv file that provide location of forest fire

-   Kepulauan_Bangka_Belitung: a ESRI shapefile showing boundary of Bangka Belitung

## Import and Wrangle Geospatial Data

```{r}
kbb_sf <- st_read(dsn="data/rawdata",
               layer = "Kepulauan_Bangka_Belitung") %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32748)
```

What does this code do:

-   `st_union()` merges all geometries in the dataset into one, so it combines multiple polygons into one multi-polygon and remove internal boundary

-   `st_zm(drop = TRUE, what = "ZM"` drops the Z (elevation) and M (measure) dimensions that are present and keep only the longitude and latitude coordinates.

-   `st_transform(crs=32748)` project geometry using WGS 84 to ensure data is in meters and suiatble to be used for analyzing regions in Indonesia

### Convert to OWIN Pbject

```{r}
kbb_owin <- as.owin(kbb_sf)
kbb_owin
```

Next, let's confirm if it's an OWIN object

```{r}
class(kbb_owin)
```

### Import and Wrangle Forest Fire Data

```{r}
fire_sf <- read_csv("data/rawdata/forestfires.csv") %>% st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% st_transform(crs = 32748)
```

What does this code do:

-   `st_as_sf(coords = c("longitude", "latitude"), crs = 4326)` converts data frame into sf object and use longitude and latitude as point coordinate and set CRS accordingly.

-   `st_transform(crs=32748)` changes coordinate reference system to be same as kbb.

Next, we'll convert data type of acq_date to numeric

```{r}
fire_sf <- fire_sf %>% 
  mutate(DayofYear = yday(acq_date)) %>%
  mutate(Month_num = month(acq_date)) %>%
  mutate(Month_fac = month(acq_date, 
                           label = TRUE, 
                           abbr = FALSE))
```

```{r}
head(fire_sf)
```

## Visualize the Fire Points

```{r}
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots()
```

Visualize Geographic Distribution of Forest Fires by Month

```{r}
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots(size = 0.1) +
tm_facets(by="Month_fac", 
            free.coords=FALSE, 
            drop.units = TRUE)
```

## Compute STKDE by Month

### Extract Forest Fires by month

First, let's remove the unwanted fields from the dataset

```{r}
fire_month <- fire_sf %>% 
  select(Month_num)
```

Next, we'll derive a ppp object

```{r}
fire_month_ppp <- as.ppp(fire_month)
fire_month_ppp
```

and check if the dataset is in the right ppp object class

```{r}
summary(fire_month_ppp)
```

and also check for duplicated point events

```{r}
any(duplicated(fire_month_ppp))
```

Next, we'll combine the kbb_owin and fire_month_ppp.

```{r}
fire_month_owin <- fire_month_ppp[kbb_owin]
summary(fire_month_owin)
```

Next, we'll visualize fire_month_owin to if it'll give the correct output

```{r}
plot(fire_month_owin)
```

## Compute Spatio-temporal KDE

We'll use `spattemp.density()` sparr library is used to compute STKDE

```{r}
st_kde <- spattemp.density(fire_month_owin)
summary(st_kde)
```

Next, we'll plot the KDE from July 2023 to December 2023

```{r}
tims <- c(7,8,9,10,11,12)
par(mfcol=c(2,3))
for(i in tims){ 
  plot(st_kde, i, 
       override.par=FALSE, 
       fix.range=TRUE, 
       main=paste("KDE at month",i))
}
```

### Compute STKDE by Day of Year

We'll select the Month_num to be included in the ppp object.

```{r}
fire_yday_ppp <- fire_sf %>% 
  select(DayofYear) %>%
  as.ppp()
```

and combine ppp object and owin object.

```{r}
fire_yday_owin <- fire_yday_ppp[kbb_owin]
summary(fire_yday_owin)
```

Next, we'll derive the KDE using spattemp.density()

```{r}
kde_yday <- spattemp.density(
  fire_yday_owin)
summary(kde_yday)
```

and plot it

```{r}
par(mar = c(2,2,2,2))
plot(kde_yday) 
```

## Compute STKDE by Day of Day: Improved Method

We can also use `BOOT.spattemp()` from sparr library to support bandwidth selection for standalone spatio-temporal density/intensity based on bootstrap estimation of the MISE, providing isotopic scalar spatial bandwidth and scalar temporal bandwidth.

```{r}
set.seed(1234)
BOOT.spattemp(fire_yday_owin) 
```

Next, we'll compute the spatio-temporal KDE using h and lambda values from previous step

```{r}
kde_yday <- spattemp.density(fire_yday_owin,
                             h = 9000,
                             lambda = 19)
summary(kde_yday)
```

and we'll plot the results

```{r}
par(mar = c(2,2,2,2))
plot(kde_yday)
```
