---
title: "Take-home Exercise 3"
author: "Stefanie Felicia"
date: "27 October, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Background

## Objective

-   In this hands-on exercise, we'll be creating a geospatial predictive model to predict HDB resale price from July to September 2024 using data from the whole year data (from July 2024 to June 2025).

The focus of this hands on exercise will be on four room HDB properties with .. as predictors.

## The Data

In this hands, on we'll be using the following datasets:

-   HDB Resale Flat Prices (from Jan-2017 onwards) taken from data.gov.sg
-   

## Import Libraries

```{r}
pacman::p_load(olsrr, corrplot, httr, tidyverse, sf, tmap, jsonlite, progress, spdep, GWmodel, SpatialML, rsample, Metrics, knitr, kableExtra, spatialRF, randomForestExplainer, ggpubr, fastDummies, performance, car, readxl)
```

Libraries used for this exercise are:

-   httr

-   tidyverse

-   sf

-   tmap

-   jsonlite

-   progress

-   spdep

-   GWmodel

-   SpatialModel

-   rsample

-   Metrics

-   knitr

-   kableExtra

-   spatialRF

## Geocode HDB

Resale Data First, we'll import the HDB resale prices data using read_csv() and filter for 4 rooms HDB data from January to September 2025.

```{r}
HDBresale <- read_csv("data/resale_price_from_jan2014.csv") %>%
  filter(flat_type == "4 ROOM") %>%
  filter(month >= "2025-01" & month <= "2025-09")
```

Next, we'll also rename street names that has "ST." into "SAINT"

```{r}
HDBresale$street_name <- gsub("ST\\.", "SAINT", 
                              HDBresale$street_name)
```

Next, we'll be combining the block and street name column into addresses to be used for geocoding

```{r}
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep = " ")
  query <- list("searchVal" = address, 
                "returnGeom" = "Y",
                "getAddrDetails" = "N", 
                "pageNum" = "1")
  
  res <- GET(base_url, query = query)
  restext <- content(res, as="text")
  
  output <- fromJSON(restext) %>%
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)    
  return(output)
  } 
```

::: callout-note
-   `"searchVal" = address`: pass the address as searchVal in the query

-   `"returnGeom" = "Y"`: send the query to OneMapSG search and return the geometry

-   set `getAddreDetails = 'N'` since don't need all the address details

-   `fromJSON(restext):` convert response from json format to text

-   `select(results.LATITUDE, results.LATITUDE):` retain latitude and longitude for output
:::

```{r}
HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0
for (i in 1:nrow(HDBresale)){
  temp_output <- geocode(HDBresale[i,4], HDBresale[i, 5])
  
HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
  
```

## Proximity Analysis

In this section, we'll count the number of geographic entities located within defined distance from each HDB properties.

Before performing proximity analysis, we need to make sure the input data need to be in sf object and they must be in the same projection coordinate system.

In our HDBresale variable, since it's still in tbl dataframe, we'll transform it into a sf object and adding in the coordinates and change the projected coordinate system.

```{r}
HDBresale_sf <- HDBresale %>%
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"), 
           crs=4326) %>%
  st_transform(crs=3414)
```

### Import CBD Data

```{r}
map <- read_rds("data/mpsz_cl.rds")
st_crs(map)
```

```{r}
cbd_points <- map %>%
  filter(REGION_N == "CENTRAL REGION")
```

```{r}
st_is_valid(cbd_points)
```

```{r}
st_crs(HDBresale_sf)
st_crs(cbd_points)
```

```{r}
cbd_dist <- st_distance(HDBresale_sf, cbd_points)

HDBresale_sf$cbd_distance <- apply(cbd_dist, 1, min) %>% as.numeric()
sum(HDBresale_sf$cbd_distance < 0) 
```

```{r}
tm_shape(map) + tm_polygons(col = "lightgrey", border.col = "white") + tm_shape(HDBresale_sf) + tm_dots(ol="red", size = 0.1)
```

### Import Hawker Center Data

```{r}
hawker <- st_read("data/HawkerCentresKML.kml") %>%
  st_transform(crs=3414)
```

```{r}
hawker_dist <- st_distance(HDBresale_sf, hawker)

HDBresale_sf <- HDBresale_sf %>%
  mutate(nearest_hawker = apply(hawker_dist, 1, min))
```

### Import Eldercare Data

```{r}
eldercare <- st_read("data/EldercareServicesKML.kml") %>%
  st_transform(crs=3414)
```

```{r}
eldercare_dist <- st_distance(HDBresale_sf, eldercare)

HDBresale_sf <- HDBresale_sf %>%
  mutate(nearest_eldercare = apply(eldercare_dist, 1, min))
```

### Import Supermarket Data

```{r}
supermarket <- st_read("data/SupermarketsKML.kml") %>%
  st_transform(crs=3414)
```

```{r}
supermarket_dist <- st_distance(HDBresale_sf, supermarket)

HDBresale_sf <- HDBresale_sf %>%
  mutate(nearest_supermarket = apply(supermarket_dist, 1, min))
```

### Import Kindergaten Data

```{r}
kindergarten <- st_read("data/Kindergartens.kml") %>%
  st_transform(crs=3414)
```

```{r}
kindergarten_within_350m <- st_is_within_distance(
  HDBresale_sf, kindergarten, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(kindergarten_350m = map_int(kindergarten_within_350m, length))
```

### Import Preschool Location Data

Next, we'll import the preschool dataset using st_read() and transform the projected coordinate system into 3414.

```{r}
preschool <- st_read("data/PreSchoolsLocation(1).kml") %>%
  st_transform(crs=3414)
```

Next, we'll count the number of preschool within 350 meter from each HDB property.

```{r}
within_350m <- st_is_within_distance(
  HDBresale_sf, preschool, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(preschool_350m = map_int(within_350m, length))
```

Import Bus Stop Location

```{r}
bus_stop <- st_read(dsn = "data/LTADataMall/", 
                  layer = "BusStop") %>%
  st_transform(crs = 3414)
```

```{r}
bus_within_350m <- st_is_within_distance(
  HDBresale_sf, bus_stop, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(bus_350m = map_int(bus_within_350m, length))
```

```{r}
HDBresale_sf$remaining_lease <- as.integer(substr(HDBresale_sf$remaining_lease, 1, 2))
```

```{r}
mrt <- read_csv("data/MRT Stations.csv")
```

```{r}
mrt_sf <- st_as_sf(mrt,
                   wkt = "geometry",  
                   crs = 4326)
```

```{r}
mrt_sf <- st_transform(mrt_sf, 3414)
```

```{r}
mrt_dist <- st_distance(HDBresale_sf, mrt_sf)

HDBresale_sf <- HDBresale_sf %>%
  mutate(nearest_mrt = apply(mrt_dist, 1, min))
```

## EDA

```{r}
ggplot(data=HDBresale_sf, 
       aes(x=`resale_price`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r}
PROX_ELDERLYCARE <- ggplot(data=HDBresale_sf, 
                           aes(x= `nearest_eldercare`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_CBD <- ggplot(data=HDBresale_sf, 
                               aes(x= `cbd_distance`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_HAWKER_MARKET <- ggplot(data=HDBresale_sf, 
                             aes(x= `nearest_hawker`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_KINDERGARTEN <- ggplot(data=HDBresale_sf, 
                            aes(x= `kindergarten_350m`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_PRESCHOOL <- ggplot(data=HDBresale_sf, 
                            aes(x= `preschool_350m`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_SUPERMARKET <- ggplot(data=HDBresale_sf, 
                            aes(x= `nearest_supermarket`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")


PROX_REMAINING_LEASE <- ggplot(data=HDBresale_sf, 
                   aes(x= `remaining_lease`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_BUS <- ggplot(data=HDBresale_sf, 
                   aes(x= `bus_350m`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

ggarrange(PROX_ELDERLYCARE, PROX_CBD, PROX_HAWKER_MARKET, PROX_KINDERGARTEN, PROX_PRESCHOOL, PROX_SUPERMARKET, PROX_REMAINING_LEASE, PROX_BUS, ncol = 5, nrow = 2)
```

Categorical Variable

```{r}
unique(HDBresale_sf$flat_model)
```

Storey Range Variable

```{r}
colnames(HDBresale_sf)
```

```{r}
nrow(HDBresale_sf)
HDBresale_sf$storey_range <- factor(HDBresale_sf$storey_range,
                         levels = c("01 TO 03", "04 TO 06", "07 TO 09", "10 TO 12",
                                    "13 TO 15", "16 TO 18", "19 TO 21", "22 TO 24",
                                    "25 TO 27", "28 TO 30", "31 TO 33", "34 TO 36",
                                    "37 TO 39", "40 TO 42", "43 TO 45", "46 TO 48",
                                    "49 TO 51"),
                         ordered = TRUE)
class(HDBresale_sf$storey_range)
```

```{r}
# drop geometry column
HDB_numeric <- st_drop_geometry(HDBresale_sf)
colnames(HDB_numeric)
```

```{r}
HDB_numeric <- st_drop_geometry(HDB_numeric)
sapply(HDB_numeric[c(6, 10:18)], class)
```

```{r}
corrplot(cor(HDB_numeric[,c(10:18)]),
         diag = FALSE, 
         order = "AOE",
         tl.pos = "td", 
         tl.cex = 0.5, 
         method = "number", 
         type = "upper")
```

Multiple Linear Regression Model

```{r}
resale_price.mlr <- lm(
  formula = resale_price ~ floor_area_sqm + remaining_lease + cbd_distance  + 
    nearest_hawker + nearest_eldercare +
    kindergarten_350m   +
    bus_350m, 
  data=HDBresale_sf)
summary(resale_price.mlr)

```

## Satisfying Requirements for Multiple Linear Regression

```{r}
check_collinearity(resale_price.mlr)
```

```{r}
vif_model <- check_collinearity(resale_price.mlr)
plot(vif_model)
```

```{r}
check_model(resale_price.mlr,
            check = "linearity")
```

```{r}
crPlot(resale_price.mlr, variable = "cbd_distance")
crPlot(resale_price.mlr, variable = "floor_area_sqm")
crPlot(resale_price.mlr, variable = "remaining_lease")
crPlot(resale_price.mlr, variable = "nearest_hawker")
crPlot(resale_price.mlr, variable = "kindergarten_350m")
crPlot(resale_price.mlr, variable = "nearest_eldercare")
crPlot(resale_price.mlr, variable = "bus_350m")
```

```{r}
sum(HDBresale_sf$floor_area_sqm < 0)      # usually none
sum(HDBresale_sf$remaining_lease < 0)     # should be >=0
sum(HDBresale_sf$cbd_distance <= 0)       # log(0) or negative → problem
sum(HDBresale_sf$nearest_hawker <= 0)     # log(0) → problem
sum(HDBresale_sf$kindergarten_350m < 0) 
```

```{r}
resale_price.mlr <- lm(
  formula = resale_price ~  sqrt(floor_area_sqm) + sqrt(remaining_lease) + log(cbd_distance + 1)  + 
    log(nearest_hawker) + nearest_eldercare +
    sqrt(kindergarten_350m)   +
    bus_350m, 
  data=HDBresale_sf)
summary(resale_price.mlr)
```

```{r}
resale_price.mlr2 <- lm(
  formula = resale_price ~  sqrt(floor_area_sqm) + sqrt(remaining_lease) + log(cbd_distance + 1)  + 
    log(nearest_hawker) + nearest_eldercare   +
    bus_350m, 
  data=HDBresale_sf)
summary(resale_price.mlr2)
```

```{r}
check_model(resale_price.mlr2,
            check = "linearity")
```

### Multricollinearity Check

## Multiple Linear Regression Model

Evaluation

## 
