---
title: "Hands-on_Ex09"
author: "Stefanie Felicia"
date: "6 November, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Getting Started

In this hands-on exercise, we'll explore how to model geographical accessibility using R's geospatial analysis packages.

## Import Libraries

```{r}
pacman::p_load(tmap, SpatialAcc, sf, 
               ggstatsplot, reshape2,
               tidyverse)
```

The libraries we'll be using in this hands-on exercise are:

-   sf: simple features in R to encode and analyze spatial vector data

-   tmap: used to draw thematic map

-   tidyverse: collection of R packages that for data manipulation and visualization

-   SpatialAcc: provide set of spatial accessibility measure from set of location (demand) to another set of locations (supply)

-   ggstatsplot: create graphics wth details from statistical test

-   reshape2: manipulate data from wide to long format

## The Data

The datasets used for this hands-on exercise are the following:

-   MP14_SUBZONE_NO_SEA_PL: URA Master Plan 2014 subzone boundary GIS data (from data.gov.sg)

-   hexagons: 250m radius hexagon GIS data (created using st_make_grid())

-   ELDERCARE: GIS data showing location of eldercare service (from data.gov.sg)

-   OD_MatrixL distance matrix in csv format with six fields:

    -   origin_id: unique if values of origin (i.e. fid of hexagon data set)

    -   destination_id: the unique id values of the destination (i.e. fid of ELDERCARE data set)

    -   entry_cost: perpendicular distance between the origins and nearest road

    -   network_cost: actual network distance from the origin and destination

    -   exit_cost: perpendicular distance between destination and nearest road

    -   total_cost: summation of entry_cost, network_cost and exit_cost

## Import Data

We'll import the MP14_SUBZONE_NO_SEA_PL, hexagons and EDLERCARE file using st_read()

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_NO_SEA_PL")
```

```{r}
hexagons <- st_read(dsn = "data/geospatial", layer = "hexagons") 
```

```{r}
eldercare <- st_read(dsn = "data/geospatial", layer = "ELDERCARE") 
```

## Data Preparation 

Next, we'll update the imported sf files into 3414 ESPG code.

```{r}
mpsz <- st_transform(mpsz, 3414)
eldercare <- st_transform(eldercare, 3414)
hexagons <- st_transform(hexagons, 3414)
```

After transforming the projection metadata, we'll verify the projected coordinate system to confirm if it had been transformed into the correct ESPG code.

```{r}
st_crs(mpsz)
```

### Clean and Update Attribute Fields of Geospatial Data

There are redundant fields in both eldercare and hexagons data table, we'll exclude the redundant fields and create new field named demand and capacity into both hexagons and eldercare sf dataframe.

```{r}
eldercare <- eldercare %>%
  select(fid, ADDRESSPOS) %>%
  mutate(capacity = 100)
```

```{r}
hexagons <- hexagons %>%
  select(fid) %>%
  mutate(demand = 100)
```

### Import Distance Matrix

Next, we'll import the OD_Matrix file using read_csv

```{r}
ODMatrix <- read_csv("data/aspatial/OD_Matrix.csv", skip = 0)
```

### Tidying Distance Matrix

Next, we'll use spread() to transform the O-D matrix from a thin format into a fat format as it is currently still in distance matrix columnwise (column = destination, row = origins).

```{r}
distmat <- ODMatrix %>%
  select(origin_id, destination_id, total_cost) %>%
  spread(destination_id, total_cost)%>%
  select(c(-c('origin_id')))
```

Since, the distance is measured in metre because SVY21 projected coordinate system is used, we'll convert the unit f measurement from metre to kilometre.

```{r}
distmat_km <- as.matrix(distmat/1000)
```

### Modelling and Visualizing Accessibility using Hansen Method

First, we'll compute Hansen's accesibility using ac().

```{r}
acc_Hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 2, 
                            family = "Hansen"))
```

Since the field name is messy, we'll rename it to accHansen

```{r}
colnames(acc_Hansen) <- "accHansen"
```

Next, we'll convert it into tibble format

```{r}
acc_Hansen <- as_tibble(acc_Hansen)
```

We'll bind the acc_Hansen with hexagon sf object

```{r}
hexagon_Hansen <- bind_cols(hexagons, acc_Hansen)
hexagon_Hansen
```

### Visualize Hansen's Accessibility

First, we'll extract the extend of hexagons sf object using st_bbox()

```{r}
mapex <- st_bbox(hexagons)
```

And we'll visualize accessibility to eldercare centre in Singapore using tmap.

```{r}
tm_shape(hexagon_Hansen,
         bbox = mapex) + 
  tm_polygons(fill = "accHansen", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: Hansen method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```

### Statistical Graphic Visualization

In this section, we'll compare the distribution of Hansen's accessibility values by URA Planning Region.

First, we'll add the planning region field to hexagon_Hansen sf

```{r}
hexagon_Hansen <- st_join(hexagon_Hansen, mpsz, 
                          join = st_intersects)
```

Next, we'll visualize the distribution using boxplot graphical method

```{r}
ggplot(data=hexagon_Hansen, 
       aes(y = log(accHansen), 
           x= REGION_N)) +
  geom_boxplot() +
  geom_point(stat="summary", 
             fun.y="mean", 
             colour ="red", 
             size=2)
```

## Model and Visualize Accessibility using KD2FCA Method

### Compute KD2SFCA's Accessibility

In this section, we'll calculates Hansen's accessibility using ac()

```{r}
acc_KD2SFCA <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            d0 = 5,
                            power = 2, 
                            family = "KD2SFCA"))

colnames(acc_KD2SFCA) <- "accKD2SFCA"
acc_KD2SFCA <- as_tibble(acc_KD2SFCA)
hexagon_KD2SFCA <- bind_cols(hexagons, acc_KD2SFCA)
```

### Visualize KD2SFCA's Accessibility

Next, we'll use tmap to visualize eldercare centre in Singapore using tmap.

```{r}
tm_shape(hexagon_KD2SFCA,
         bbox = mapex) + 
  tm_polygons(fill = "accKD2SFCA", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: KD2SFCA method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```

### Statistical Graphic Visualization

Next, we'll compare the distribution of KD2CFA accessibility values by URA Planning Region

First, we'll add planning region field to hexagon_KD2FCA sf dataframe

```{r}
hexagon_KD2SFCA <- st_join(hexagon_KD2SFCA, mpsz, 
                          join = st_intersects)
```

Next., we'll use ggplot() to plot the distribution using boxplot.

```{r}
ggplot(data=hexagon_KD2SFCA, 
       aes(y = accKD2SFCA, 
           x= REGION_N)) +
  geom_boxplot() +
  geom_point(stat="summary", 
             fun.y="mean", 
             colour ="red", 
             size=2)
```

## Model and Visualize Accessibility using Spatial Accessibility Measure Method

### Compute SAM Accesibility

In this section, we'll compute SAM accesibility using ac(), but with `family = "SAM"`.

```{r}
acc_SAM <- data.frame(ac(hexagons$demand,
                         eldercare$capacity,
                         distmat_km, 
                         d0 = 5,
                         power = 2, 
                         family = "SAM"))

colnames(acc_SAM) <- "accSAM"
acc_SAM <- as_tibble(acc_SAM)
hexagon_SAM <- bind_cols(hexagons, acc_SAM)
```

### Visualize SAM's Accessibility

We'll visualize the eldercare accessibility using SAM accessibility method.

```{r}
tm_shape(hexagon_SAM,
         bbox = mapex) + 
  tm_polygons(fill = "accSAM", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: SAM method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```

### Statistical Graphic Visualization

Next, we'll compare the distribution of SAM accessibility values by URA Planning Region.

First, we'll add planning region field to hexagon_SAM sf

```{r}
hexagon_SAM <- st_join(hexagon_SAM, mpsz, 
                       join = st_intersects)
```

We'll use ggplot() to plot distribution using boxplot.

```{r}
ggplot(data=hexagon_SAM, 
       aes(y = accSAM, 
           x= REGION_N)) +
  geom_boxplot() +
  geom_point(stat="summary", 
             fun.y="mean", 
             colour ="red", 
             size=2)
```
