---
title: "In-class Exercise 5b"
author: "Stefanie Felicia"
date: "27 September, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Import libraries

```{r}
pacman::p_load(sf,sfdep,tmap,tidyverse)
```

## The Data

### Importing Hunan table

```{r}
 hunan <- st_read(dsn = "data/geospatial", layer = "Hunan")
names(hunan)
```

### Importing GDPPC table

```{r}
GDPPC <- read_csv("data/aspatial/Hunan_GDPPC.csv")
names(GDPPC)
```

### Joining Hunan and Hunan_2012

```{r}
GDPPC_st <- spacetime(GDPPC, hunan, .loc_col = "County",
.time_col = "Year")
```

```{r}
is_spacetime_cube(GDPPC_st)
```

## Derive Spatial Weights

```{r}
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
  mutate(nb = include_self(
    st_contiguity(geometry)),
    wt = st_inverse_distance(nb, 
                             geometry, 
                             scale = 1,
                             alpha = 1),
    .before = 1) %>%
  set_nbs("nb") %>%
  set_wts("wt")
```

## Computing Gi

```{r}
gi_stars <- GDPPC_nb %>% 
  group_by(Year) %>% 
  mutate(gi_star = local_gstar_perm(
    GDPPC, nb, wt)) %>% 
  tidyr::unnest(gi_star)
```

## Mann-Kendall Test on Gi

```{r}
cbg <- gi_stars %>% 
  ungroup() %>% 
  filter(County == "Changsha") |> 
  select(County, Year, gi_star)
```

## Plot result using ggplot2

```{r}
ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()
```

## Print the Test Report

```{r}
cbg %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>% 
  tidyr::unnest_wider(mk)
```

## Mann-Kendall Test Data frame

```{r}
ehsa <- gi_stars %>%
  group_by(County) %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>%
  tidyr::unnest_wider(mk)
head(ehsa)
```

We can also sort to show the significance of emerging hot or cold spots

```{r}
emerging <- ehsa %>% 
  arrange(sl, abs(tau)) %>% 
  slice(1:10)
head(emerging)
```

## Perform Emerging Hotspot Analysis

```{r}
ehsa <- emerging_hotspot_analysis(
  x = GDPPC_st, 
  .var = "GDPPC", 
  k = 1, 
  nsim = 99
)
```

## Visualize the distributions of EHSA classes

```{r}
ggplot(data = ehsa,
       aes(x = classification)) +
  geom_bar()
```

## Visualizing EHSA

```{r}
hunan_ehsa <- hunan %>%
  left_join(ehsa,
            by = join_by(County == location))
```

```{r}
ehsa_sig <- hunan_ehsa  %>%
  filter(p_value < 0.05)
tmap_mode("plot")
tm_shape(hunan_ehsa) +
  tm_polygons() +
  tm_borders(alpha = 0.5) +
tm_shape(ehsa_sig) +
  tm_fill("classification") + 
  tm_borders(alpha = 0.4)
```
