---
title: "In-class_Ex09"
format: html
---

```{r}
pacman::p_load(SpatialAcc, sf, tidyverse, 
               tmap, ggstatsplot)
```

## Count Number of Points within a Distance

In this section, we'll count the number of point features of CHAS clinic within 1 km of another point feature (eldercare centre)

```{r}
eldercare <- st_read(dsn = "data/geospatial",
                     layer = "ELDERCARE") %>%
  st_transform(crs = 3414)
```

```{r}
CHAS <- st_read("data/geospatial/CHASClinics.kml") %>%
  st_transform(crs = 3414)
```

```{r}
buffer_1km <- st_buffer(eldercare, 
                        dist = 1000)
```

```{r}
tmap_mode("view")
tm_shape(buffer_1km) +
  tm_polygons() +
tm_shape(CHAS) +
  tm_dots()
```

```{r}
buffer_1km$pts_count <- lengths(
  st_intersects(buffer_1km, CHAS))
```

## Import Data

```{r}
mpsz <- st_read(dsn = "data/geospatial",
                layer = "MP14_SUBZONE_NO_SEA_PL") %>%
  st_transform(crs = 3414)

hexagons <- st_read(dsn = "data/geospatial", layer = "hexagons") %>%
  st_transform(crs = 3414)

eldercare <- st_read(dsn = "data/geospatial",
                     layer = "ELDERCARE") %>%
  st_transform(csr = 3414)
```

```{r}
ODMatrix <- read_csv("data/aspatial/OD_Matrix.csv", 
                     skip = 0)
```

## Data Cleaning and Update Attribute

```{r}
eldercare <- eldercare %>%
  select(fid, ADDRESSPOS) %>%
  mutate(capacity = 100)
```

```{r}
hexagons <- hexagons %>%
  select(fid) %>%
  mutate(demand = 100)
```

```{r}
distmat <- ODMatrix %>%
  select(origin_id, destination_id, total_cost) %>%
  spread(destination_id, total_cost)%>%
  select(c(-c('origin_id')))
```

```{r}
distmat_km <- as.matrix(distmat/1000)
distmat_km
```

## Compute Handsen's Accessibility

```{r}
acc_Hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 2, 
                            family = "Hansen"))
```

## Visualize Accessibility

```{r}
colnames(acc_Hansen) <- "accHansen"

acc_Hansen <- as_tibble(acc_Hansen)

hexagon_Hansen <- bind_cols(hexagons, acc_Hansen)
```

```{r}
mapex <- st_bbox(hexagons)

tmap_mode("plot")
tm_shape(hexagon_Hansen, bbox = st_bbox(hexagons)) +
  tm_polygons(
    fill = "accHansen",
    fill.scale = tm_scale_intervals(style = "quantile", n = 10),
    fill.legend = tm_legend(title = "Accessibility (Hansen)"),
    border.col = "black"
  ) +
tm_shape(eldercare) +
  tm_symbols(size = 0.1) +
  tm_compass(type = "8star", size = 2) +
  tm_grid(lwd = 0.1, alpha = 0.5) +
  tm_title("Accessibility to eldercare: Hansen method",
           position = c("center", "top"),
           size = 1.2) +
  tm_legend(position = c("right", "top")) +
  tm_scalebar() +   # <— move to the very end
  tm_layout(frame = TRUE)

```

::: {.callout-note}
-   Most of Singapore are not accessible for eldercare, they need to travel far to be able to get to the nearest eldercare. There are only a few points in other areas and in central area, which indicate that there are more eldercare centre at those areas.

-   Another interpretation could be that there are small facilities eldercares centre nearby, so there are limited spots, so that there isn't many elderly that can take that much elderly.
:::

## Statistical Graph

```{r}
hexagon_Hansen <- st_join(hexagon_Hansen, mpsz, 
                          join = st_intersects)
```

```{r}
ggbetweenstats(
  data = hexagon_Hansen,
  x = REGION_N,
  y = accHansen,
  type = "p")
```

::: {.callout-note}
-   As seen in the observation before, the central region has the highest number of eldercare centre, with north region having the least eldercare centre.
:::
